Does the COMPAS Needle Always Point Towards Equity? Finding Fairness in the COMPAS Risk Assessment Algorithm: A Case Study

### 2022-04-19 Paper_Draft2/Draft2.Rmd
hyperlink places that use COMPAS instead of reference

author:
  - name: Amrita Acharya
    affil: 1
  - name: Dianne Caravela
    affil: 1
  - name: Eunice Kim
    affil: 1
  - name: Emma Kornberg
    affil: 1
  - name: Elisabeth Nesmith
    affil: 1
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
    email: aarcharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu

@@

The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [Equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism. It generates a decile score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine]. Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the New York [@NYDOCCS2019], Massachusetts, Michigan, California [@Jackson2020Setting], and Wisconsin [@WisconsinDOC] Departments of Corrections. 
Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
@angwin2016machine. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].

@@

The statistical parity difference is -0.14. This indicates that there is a large difference between white and Black defendants regarding whether or not they recidivate. The algorithm unfairly benefits white defendants over Black defendants.

Disparate impact ratio is 0.47. The ratio of white defendants predicted to not recidivate to the Black defendants predicted to not recidivate is 0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the algorithm unfairly benefits white defendants.

>>>>

author:
  - name: Amrita Acharya
    affil: 1
  - name: Dianne Caravela
    affil: 1
  - name: Eunice Kim
    affil: 1
  - name: Emma Kornberg
    affil: 1
  - name: Elisabeth Nesmith
    affil: 1
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
    email: aacharya@smith.edu, dcaravela@smith.edu, ekim89@smith.edu, ekornberg@smith.edu, enesmith@smith.edu

@@

The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private, for-profit company Northpointe (now known by its parent company [equivant](https://www.equivant.com/faq/)), to predict defendants’ risk of recidivism. It generates a decile score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine]. Jurisdictions across the United States use the COMPAS risk assessment instrument, including but not limited to the [New York](https://doccs.ny.gov/system/files/documents/2020/11/8500.pdf), [Massachusetts](https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4), [Michigan](https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4), [California](https://hdsr.mitpress.mit.edu/pub/hzwo7ax4/release/4), and [Wisconsin](https://doc.wi.gov/Pages/AboutDOC/COMPAS.aspx) Departments of Corrections.

Due to the proprietary nature of the COMPAS algorithm, it is unknown how exactly these recidivism risk scores are calculated; however, a sample COMPAS Risk Assessment Survey has been made publicly available, revealing the algorithm’s input information. This survey has been widely critiqued for using proxy variables for race that do not explicitly factor in a defendant’s race but heavily imply it, allowing Northpointe to claim that their algorithm is free of racial bias. For example, the COMPAS risk assessment survey asks screeners to speculate if a defendant might be affiliated with a gang. It also asks if a defendant has any friends or family members who have been crime victims [@Angwin2016Sample]. Although these questions do not directly ask about race, they do not take into account the pervasive nature of systemic racism that infiltrates every aspect of the lives of marginalized people, thereby indirectly asking about race. 
@angwin2016machine. published a 2016 piece in the news publication ProPublica that analyzes the methods and algorithms used by Northpointe in their COMPAS risk score assessment algorithm and uncovers the racial biases in defendants’ scores [@angwin2016machine]. Their research finds that “the algorithm [is] somewhat more accurate than a coin flip,” a worrisome level of accuracy given the potential impact its determinations may have on real people’s lives. @angwin2016machine specifically investigate the distribution of COMPAS scores by decile among Black and white defendants. They write: “The analysis also [shows] that even when controlling for prior crimes, future recidivism, age, and gender, Black defendants [are] 45 percent more likely to be assigned higher risk scores than white defendants” [@larson2016we]. After examining the fairness metric statistical parity difference, @angwin2016machine conclude that the algorithm is racially biased [@larson2016we].

@@

The statistical parity difference is -0.14. This indicates that there is a large difference between white and Black defendants regarding whether or not they recidivate. The algorithm unfairly benefits white defendants over Black defendants. 
Disparate impact ratio is 0.47. The ratio of white defendants predicted to not recidivate to the Black defendants predicted to not recidivate is 0.47. A ratio between 0.8 and 1.25 is considered fair, therefore the algorithm unfairly benefits white defendants.

### 2022-04-19 Paper_Draft2/Draft2.tex
hyperlink places that use COMPAS instead of reference

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA 01063; \\
$^{2}$ \quad ; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}}\\
$^{3}$ \quad ; \href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}}\\
$^{4}$ \quad ; \href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}}\\
$^{5}$ \quad ; \href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}}\\
$^{6}$ \quad ; \href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: }

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A variety of disciplines use risk assessment instruments to help humans
make data-driven decisions. Northpointe, a software company, created an
algorithmic risk assessment instrument known as the Correctional
Offender Management Profiling for Alternative Sanctions (COMPAS). COMPAS
uses various behavioral and psychological metrics related to recidivism
to assist justice systems in assessing a defendant's potential
recidivism risk. Angwin et al.~published a ProPublica article in which
they conclude that the racial biases in the criminal justice system are
reflected in the COMPAS recidivism risk scores. In response, Dieterich
et al.~published a rebuttal on behalf of Northpointe defending the
COMPAS algorithm and refuting Angwin et al.'s allegation of racial bias.
Using a human rights framework adopted from the organizations Women at
the Table and AI Fairness 360, we use debiasing algorithms and fairness
metrics to analyze the argument between Northpointe and ProPublica and
determine whether and to what extent there is racial bias in the COMPAS
algorithm.}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
Departments of Corrections. Due to the proprietary nature of the COMPAS
algorithm, it is unknown how exactly these recidivism risk scores are
calculated; however, a sample COMPAS Risk Assessment Survey has been
made publicly available, revealing the algorithm's input information.
This survey has been widely critiqued for using proxy variables for race
that do not explicitly factor in a defendant's race but heavily imply
it, allowing Northpointe to claim that their algorithm is free of racial
bias. For example, the COMPAS risk assessment survey asks screeners to
speculate if a defendant might be affiliated with a gang. It also asks
if a defendant has any friends or family members who have been crime
victims \citep{Angwin2016Sample}. Although these questions do not
directly ask about race, they do not take into account the pervasive
nature of systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race. Angwin et
al.~published a 2016 piece in the news publication ProPublica that
analyzes the methods and algorithms used by Northpointe in their COMPAS
risk score assessment algorithm and uncovers the racial biases in
defendants' scores \citep{angwin2016machine}. Their research finds that
``the algorithm {[}is{]} somewhat more accurate than a coin flip,'' a
worrisome level of accuracy given the potential impact its
determinations may have on real people's lives. Angwin et
al.~specifically investigate the distribution of COMPAS scores by decile
among Black and white defendants. They write: ``The analysis also
{[}shows{]} that even when controlling for prior crimes, future
recidivism, age, and gender, Black defendants {[}are{]} 45 percent more
likely to be assigned higher risk scores than white defendants''
\citep{larson2016we}. After examining the fairness metric statistical
parity difference, Angwin et al.~conclude that the algorithm is racially
biased \citep{larson2016we}.

Dieterich et al., on the behalf of Northpointe, deny the allegations of
racial bias and offer their own analyses based on different fairness
metrics in rebuttal \citep{equivant_response_2016}. Angwin et
al.~maintain that there are biases in the outcome values, protected
attributes, and covariates during Dieterich et al.'s data processing
phase. ProPublica collaborators Larson et al.~account for these biases
in their analyses. In their response, Dieterich et al.~highlight that
Angwin et al.~did not account for base rates of recidivism in their
analysis, which are important to understand initial percentages without
the presence of other information.

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple data sets, including the
COMPAS data set that accompanied the Angwin et al.~piece.

>>>>

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Statistical and Data Sciences Smith College Northampton, MA
01063; \href{mailto:aacharya@smith.edu}{\nolinkurl{aacharya@smith.edu}},
\href{mailto:dcaravela@smith.edu}{\nolinkurl{dcaravela@smith.edu}},
\href{mailto:ekim89@smith.edu}{\nolinkurl{ekim89@smith.edu}},
\href{mailto:ekornberg@smith.edu}{\nolinkurl{ekornberg@smith.edu}},
\href{mailto:enesmith@smith.edu}{\nolinkurl{enesmith@smith.edu}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:bbaumer@smith.edu}{\nolinkurl{bbaumer@smith.edu}}}

@@

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A variety of disciplines use risk assessment instruments to
help humans make data-driven decisions. Northpointe, a software company,
created an algorithmic risk assessment instrument known as the
Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS). COMPAS uses various behavioral and psychological metrics
related to recidivism to assist justice systems in assessing a
defendant's potential recidivism risk. \citet{angwin2016machine}
published a ProPublica article in which they conclude that the racial
biases in the criminal justice system are reflected in the COMPAS
recidivism risk scores. In response, \citet{equivant_response_2016}
published a rebuttal on behalf of Northpointe defending the COMPAS
algorithm and refuting \citet{angwin2016machine}'s allegation of racial
bias. Using a human rights framework adopted from the organizations
Women at the Table and AI Fairness 360, we use debiasing algorithms and
fairness metrics to analyze the argument between Northpointe and
ProPublica and determine whether and to what extent there is racial bias
in the COMPAS algorithm.}

@@

The Correctional Offender Management Profiling for Alternative Sanctions
(COMPAS) algorithm was created by the private, for-profit company
Northpointe (now known by its parent company
\href{https://www.equivant.com/faq/}{Equivant}), to predict defendants'
risk of recidivism. It generates a decile score that classifies
defendants' risk of recidivism as either low, medium, or high
\citep{angwin2016machine}. Jurisdictions across the United States use
the COMPAS risk assessment instrument, including but not limited to the
New York \citep{NYDOCCS2019}, Massachusetts, Michigan, California
\citep{Jackson2020Setting}, and Wisconsin \citep{WisconsinDOC}
Departments of Corrections.

Due to the proprietary nature of the COMPAS algorithm, it is unknown how
exactly these recidivism risk scores are calculated; however, a sample
COMPAS Risk Assessment Survey has been made publicly available,
revealing the algorithm's input information. This survey has been widely
critiqued for using proxy variables for race that do not explicitly
factor in a defendant's race but heavily imply it, allowing Northpointe
to claim that their algorithm is free of racial bias. For example, the
COMPAS risk assessment survey asks screeners to speculate if a defendant
might be affiliated with a gang. It also asks if a defendant has any
friends or family members who have been crime victims
\citep{Angwin2016Sample}. Although these questions do not directly ask
about race, they do not take into account the pervasive nature of
systemic racism that infiltrates every aspect of the lives of
marginalized people, thereby indirectly asking about race.
\citet{angwin2016machine}. published a 2016 piece in the news
publication ProPublica that analyzes the methods and algorithms used by
Northpointe in their COMPAS risk score assessment algorithm and uncovers
the racial biases in defendants' scores \citep{angwin2016machine}. Their
research finds that ``the algorithm {[}is{]} somewhat more accurate than
a coin flip,'' a worrisome level of accuracy given the potential impact
its determinations may have on real people's lives.
\citet{angwin2016machine} specifically investigate the distribution of
COMPAS scores by decile among Black and white defendants. They write:
``The analysis also {[}shows{]} that even when controlling for prior
crimes, future recidivism, age, and gender, Black defendants {[}are{]}
45 percent more likely to be assigned higher risk scores than white
defendants'' \citep{larson2016we}. After examining the fairness metric
statistical parity difference, \citet{angwin2016machine} conclude that
the algorithm is racially biased \citep{larson2016we}.

\citet{equivant_response_2016}, on the behalf of Northpointe, deny the
allegations of racial bias and offer their own analyses based on
different fairness metrics in rebuttal \citep{equivant_response_2016}.
\citet{angwin2016machine} maintain that there are biases in the outcome
values, protected attributes, and covariates during
\citet{equivant_response_2016}'s data processing phase. ProPublica
collaborators \citet{larson2016we} account for these biases in their
analyses. In their response, \citet{equivant_response_2016} highlight
that \citet{angwin2016machine} did not account for base rates of
recidivism in their analysis, which are important to understand initial
percentages without the presence of other information.

@@

AI Fairness 360 (AIF360) is an open-source Python toolkit that seeks to
``to help facilitate the transition of fairness research algorithms to
use in an industrial setting and to provide a common framework for
fairness researchers to share and evaluate algorithms''
\citep{aif360-oct-2018}. It contains multiple data sets, including the
COMPAS data set that accompanied the \citet{angwin2016machine} piece.
